---
title: "data"
format: html
---

 
## Federally Recognized Indigenous Nations

The following visualizations draw on U.S. Census Bureau data to map 
federally recognized Indigenous nations across the United States. This 
data is directly relevant to my research on Indigenous sovereignty and 
the ongoing struggle for federal recognition. I don't want to give the impression that this is the totality of Indigenous land, this is only the top 50 nations by landmass. The intent was to show mapping from Digital History Methods, and to invite people to think about their own proximity to Indigenous people, who are often made to feel far away, and long ago. This map shows that there are Indigenous people dispersed all over the United States. and this is just the sample of the top 50 by sq mileage. Note the void of Ohio Tennessee, Kentucky, Pennsylvania etc. I do not know the whole story of how. But it is notable that these states have no federally recognized nations. I wonder if there is potential to advocate for recognition of groups within those states, although removal in the 19th century seems likely. 

*Data source: U.S. Census Bureau via the tigris R package.*

---

## Interactive Map of Indigenous Nation Lands

```{r}
#| echo: false
#| message: false
#| warning: false

# Load the packages we need
library(tigris)    # connects to Census Bureau geographic data
library(leaflet)   # builds interactive maps for the web
library(dplyr)     # tools for manipulating and reshaping data

# I am getting this weird percentile thing happening on the quarto preview. I am going to suppress in different spots to see if it goes away and works. 

options(tigris_use_cache = TRUE) 

# Download Indigenous nation boundary data from the Census Bureau
# cb = TRUE means use a simplified/compressed version (loads faster)
nations <- native_areas(cb = TRUE)

# Build the interactive map
leaflet(nations) |>                        # start a leaflet map using the nations data
  addProviderTiles("CartoDB.Positron") |>  # add a clean minimal background map
  addPolygons(                             # draw the nation boundary shapes
    fillColor = "brown",                   # fill color inside the boundaries
    fillOpacity = 0.5,                     # how transparent the fill is (0=invisible, 1=solid)
    color = "blue",                        # color of the boundary lines
    weight = 2,                            # thickness of the boundary lines
    popup = ~NAME,                         # when clicked, show the nation name
    highlight = highlightOptions(          # what happens when you hover over a nation
      weight = 3,                          # make the border thicker on hover
      fillOpacity = 0.8,                   # make the fill more solid on hover
      bringToFront = TRUE                  # bring hovered shape in front of others
    )
  ) |>
  addControl(                              # add a text label on the map
    "<b>Federally Recognized Indigenous Nation Lands</b><br>Click any area for nation name",
    position = "topright"                  # place the label in the top right corner
  )
```

---

## Fifty Largest Indigenous Nation Land Areas

```{r}
#| echo: false
#| message: false
#| warning: false

# Load packages needed for this chunk
library(tigris)    # connects to Census Bureau geographic data
library(ggplot2)   # builds charts and graphs
library(dplyr)     # tools for manipulating and reshaping data

# options(tigris_use_cache = TRUE) 

# Re-download the nations data for this chunk
# Each chunk runs independently so need the data here too 
nations <- native_areas(cb = TRUE)

# Prepare the data for the bar chart
nations_df <- nations |>                   # start with the nations data
  as.data.frame() |>                       # convert from geographic format to a plain table
  mutate(acres = ALAND / 4047) |>          # create new column converting square meters to acres
  arrange(desc(acres)) |>                  # sort from largest to smallest by acreage
  slice(1:50)                              # keep only the top 50 rows approx 10% of nations

# Build the bar chart
ggplot(nations_df,                         # tell ggplot which dataset to use
  aes(x = reorder(NAME, acres),            # x axis = nation names ordered by acreage
      y = acres / 1000000)) +              # y axis = acreage divided by 1 million for readability
  geom_col(fill = "red",                   # draw bars filled in red
           alpha = 0.85) +                 # slight transparency on the bars
  coord_flip() +                           # flip so nation names run horizontally
  labs(                                    # set all the text labels
    title = "Fifty Largest Federally Recognized Indigenous Nation Land Areas",
    subtitle = "U.S. Census Bureau data",
    x = NULL,                              # no label needed on x axis
    y = "Land Area (millions of acres)",
    caption = "Source: U.S. Census Bureau, tigris R package"
  ) +
  theme_minimal() +                        # visual style
  theme(
    plot.title = element_text(face = "bold"),  # make the title bold
    axis.text.y = element_text(size = 8)       # make nation name text smaller so it fits
  )
```